High-order cognition is supported by information-rich but compressible brain activity patterns

### 2023-03-15 paper/main.tex
updating main text

\newcommand{\synthThirds}{S1}
\newcommand{\accuracyNetworks}{S2}
\newcommand{\componentsNetworks}{S3}
\newcommand{\componentBrains}{S4}
\newcommand{\topTerms}{S5}
\newcommand{\neurosynthFull}{S6}

@@

We take as a given that brain activity is highly flexible: our brains can
exhibit nearly infinite activity patterns. This flexibility implies that our
brains activity patterns are highly information rich. However, brain activity
patterns are also highly structured. For example, full-brain correlation
matrices are stable within~\citep{FinnEtal15, FinnEtal17, GratEtal18} and
across~\citep{YeoEtal11, GlerEtal12, GratEtal18, ColeEtal14} individuals. This
stability suggests that our brains' activity patterns are at least partially
constrained, for example by anatomical, external, or internal factors.
Constraints on brain activity that limit its flexibility decrease
expressiveness (i.e., its information rate). However, constraints on brain
activity also increase its robustness to noise (e.g., ``missing'' or corrupted
signals may be partially recovered). For example, recent work has shown that
full-brain activity patterns may be reliably recovered from only a relatively
small number of implanted electrodes~\citep{OwenEtal20, ScanEtal21}. This
robustness property suggests that the relevant signal (e.g., underlying factors
that have some influence over brain activity patterns) are compressible.

@@

We used a previously collected neuroimaging dataset to estimate the extent to
which each of these two possibilities might hold. The dataset we examined
comprised functional magnetic resonance imaging (fMRI) data collected as
participants listened to an audio recording of a 10-minute story, temporally
scrambled recordings of the story, or underwent a resting state
scan~\citep{SimoEtal16}. Each of these experimental conditions evokes different
depths of cognitive processing~\citep{SimoEtal16,LernEtal11,
HassEtal08,OwenEtal21}. We used across-participant classifiers to decode
listening times in each condition, as a proxy for how ``informative'' the
task-specific activity patterns were~\citep{SimoChan20}. We also use principle
components analysis to generate lower-dimensional representations of the
activity patterns. We then repeated the classification analyses after
preserving different numbers of components and examined how classification
accuracy changed across the different experimental conditions.

@@

To evaluate the relation between informativeness and compressibiilty for brain
activity from each experimental condition, we trained a series of
across-participant temporal decoders on compressed representations of the data.
Figure~\ref{fig:inflection}A displays the decoding accuracy as a function of
the number of principal components used to represent the data. Several patterns
were revealed by the analysis. First, in general (i.e., across experimental
conditions), decoding accuracy improves as the number of components increases.
However, decoding accuracy peaked at higher levels for experimental conditions
that exposed participants to cognitively richer stimuli. The peak decoding
accuracy was highest for the ``intact'' condition (versus paragraph: $t(99) =
35.205, p < 0.001)$; versus word: $t(99) = 43.172, p < 0.001)$; versus rest:
$t(99) = 81.361, p < 0.001)$), next highest for the ``paragraph'' condition
(versus word: $t(99) = 6.243, p < 0.001)$; versus rest: $t(99) = 50.748, p <
0.001)$), and next highest for the ``word'' condition (versus rest: $t(99) =
48.791, p < 0.001)$). This ordering implies that cognitively richer conditions
evoke more stable brain activity patterns across people.

The cognitively richer conditions also displayed steeper initial slopes. For
example, the intact condition decoders reached an arbitrarily chosen threshold
of 5\% accuracy using fewer components than the paragraph condition decoders
($t(99) = -7.429, p < 0.001)$) or word condition decoders ($t(99) = -7.300, p <
0.001)$), and decoding accuracy never exceeded 5\% for the rest condition. This
suggests that brain activity patterns evoked by cognitively richer conditions
are more compressible, such that representing the data using the same number of
principal components provides more information to the temporal decoders
(Fig.~\ref{fig:inflection}B).

In every experimental condition, decoding accuracy appeared to asymptote (i.e.,
hit an upper limit) beyond some characteristic number of components that
differed across conditions. To quantify the ``inflection points'' at which the
decoding curves in Figure~\ref{fig:inflection}A flattened out, we fit a sigmoid
function to the average decoding curve for each condition. We defined the
inflection point for each condition as the point on the fitted sigmoid where
the second derivative was both positive and less than a threshold value of
0.0001 (i.e., approaching 0 from the right). These inflection points reflect a
``balance'' between higher decoding accuracy (which tends to be better when
more components are used) and compression (which is better for fewer
components). Plotting each condition's inflection point
(Fig.~\ref{fig:inflection}D) reveals that both the number of components and the
decoding accuracy at each inflection point increase systematically across
conditions in proportion to cognitive richness.

@@

\label{fig:inflection-thirds}

@@

If informativeness (to the temporal decoders) and compressibility vary with the
cognitive richness of the stimulus, might these measures also vary over time
\textit{within} a given condition? For example, participants in the intact
condition might process the ongoing story more deeply later on in the story
(compared with earlier in the story) given the additional narrative background
and context they had been exposed to by that point. To examine this
possibility, we divided each condition into three successive time segments. We
computed decoding curves (Fig.~\ref{fig:inflection-thirds}A) and inflection
points (Fig.~\ref{fig:inflection-thirds}B) for each segment and condition. We
found that, in the two most cognitively rich conditions (intact and paragraph),
both decoding accuracy and compressibility, as reflected by the change in
decoding curves, increased with listening time (e.g., at the annotated
reference point of $k = 20$ components: intact: $t(99) = 7.915, p < 0.001)$;
paragraph: $t(99) = 2.354, p = 0.021)$). These changes may reflect an increase
in comprehension or depth of processing with listening time. In contrast, the
decoding accuracy and compressibility \textit{decreased} with listening time in
the word condition ($t(99) = -10.747, p < 0.001)$) and rest condition ($t(99) =
-22.081, p < 0.001)$). This might reflect the depletion of attentional
resources in the less-engaging word and rest conditions.

@@

We also wondered how informativeness and compressibility in the different
experimental conditions might vary across brain networks. We used a network
parcellation identified by \cite{YeoEtal11} to segment the brain into seven
distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Fig.~\ref{fig:networks}A):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy (for the
intact condition) tended to increase (Fig.~\ref{fig:networks}B). This suggests
that higher-order networks may carry more content-relevant or stimulus-driven
``information.'' We found no clear trends in the numbers of components at each
network's inflection point across networks or conditions
(Fig.~\ref{fig:networks}C).

@@

\caption{\textbf{Neurosynth topic weightings by component.} We used a reverse
inference procedure (see \textit{Reverse inference}) to estimate the
correspondences between brain images and a set of 80 topics derived from the
Neurosynth database of neuroimaging articles~\citep{RubiEtal17}. \textbf{A.
Topic correlations by component.} For each of the top five highest-weighted
principal components (columns) derived from each experimental condition
(colors), the radar plots display the correlations between the component images
and the per-topic images derived for each topic (topic identities are denoted
by the blue dots; a legend for the topic labels is in the lower right of the
figure). The top-weighted terms for each topic may be found in Table~\topics.
\textbf{B. Topic correlations averaged across components.} The radar plot is in
the same format as the plots in Panel A, but here we display the per-condition
average correlations across the top five components (for each condition)
reflected in Panel A. \textbf{C. Component images.} Each plot diplays a right
sagittal view of a glass brain projection of the top five principal components
(columns) for each experimental conditions (rows). Additional projections for
each component may be found in Figure~\componentBrains.}

@@

In addition to examining different networks in isolation, we wondered about the
general structure of the full-brain (i.e., potentially multi-network) activity
patterns reflected by different principal components across different
experimental conditions. Figure~\ref{fig:neurosynth-pca} displays inflated
brain maps of the top five highest-weighted components, for each experimental
condition. We also used Neurosynth~\citep{RubiEtal17} to identify, for each
component, the top five terms associated with each map (see \textit{Reverse
inference}). We noticed (by inspection) several common themes across the sets
of terms associated with each component and condition. Memory-related
components included terms like ``middle temporal,'' ``memory retrieval,'' and
``working memory.'' Sensory processing related components included terms like
``primary visual,'' ``auditory cortex,'' ``v1,'' and so on. Other components
were associated with sensory integration (e.g., ``thalamic,'' ``cingulate,''
etc.), sentence comprehension (e.g., ``sentence,'' ``comprehension''), emotion
and valence (e.g., ``amygdala,'' ``valance''), or the default mode network
(e.g., ``default mode'').  The components we identified were relatively stable
across story segments (Fig.~\synthThirds).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/components_neurosynth_summary}

\caption{\textbf{Summary of functions associated with top-weighted components
by condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and several core cognitive function classifications is
displayed in the lower right of the figure. Table~\topics~provides a list of
each topic's top-weighted terms, along with each topic's manually labeled
cognitive classification. A full list of the topics most highly associated with
each component may be found in Figure~\topTerms. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heatmaps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heatmaps display the correlations between the brain maps
(Fig.~\componentBrains) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition components, by
component.} Each sub-panels network plot summarizes the pattern of correlations
between the $n$\textsuperscript{th} top-weighted principal components
(sub-panel) for each experimental condition (gray-outlined dots). The line
thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heatmaps in Panel C). Also see
Fig.~\neurosynthFull~for additional information.}

>>>>

\newcommand{\varExplained}{S1}
\newcommand{\topTerms}{S2}
\newcommand{\componentBrains}{S3}
\newcommand{\topicContrasts}{S4}
\newcommand{\neurosynthFull}{S5}

@@

We take as a given that brain activity is highly flexible: our brains can
exhibit nearly infinite varieties of activity patterns. This flexibility implies that our
brains activity patterns are highly information rich. However, brain activity
patterns are also highly structured. For example, full-brain correlation
matrices are stable within~\citep{FinnEtal15, FinnEtal17, GratEtal18} and
across~\citep{YeoEtal11, GlerEtal12, GratEtal18, ColeEtal14} individuals. This
stability suggests that our brains' activity patterns are at least partially
constrained, for example by anatomical, external, or internal factors.
Constraints on brain activity that limit its flexibility decrease
expressiveness (i.e., its information rate). However, constraints on brain
activity also increase its robustness to noise (e.g., ``missing'' or corrupted
signals may be partially recovered). For example, recent work has shown that
full-brain activity patterns may be reliably recovered from only a relatively
small number of implanted electrodes~\citep{OwenEtal20, ScanEtal21}. This
robustness property suggests that the relevant signal (e.g., underlying factors
that have some influence over brain activity patterns) are compressible.

@@

We used a previously collected neuroimaging dataset to estimate the extent to
which each of these two possibilities might hold. The dataset we examined
comprised functional magnetic resonance imaging (fMRI) data collected as
participants listened to an audio recording of a 10-minute story, temporally
scrambled recordings of the story, or underwent a resting state
scan~\citep{SimoEtal16}. Each of these experimental conditions evokes different
depths of cognitive processing~\citep{SimoEtal16,LernEtal11,
HassEtal08,OwenEtal21}. We used across-participant classifiers to decode
listening times in each condition, as a proxy for how ``informative'' the
task-specific activity patterns were~\citep{SimoChan20}. We also used principle
components analysis to generate lower-dimensional representations of the
activity patterns. We then repeated the classification analyses after
preserving different numbers of components and examined how classification
accuracy changed across the different experimental conditions.

@@

To evaluate the relation between informativeness and compressibiilty for brain
activity from each experimental condition, we trained a series of
across-participant temporal decoders on compressed representations of the data.
Figure~\ref{fig:inflection}A displays the decoding accuracy as a function of
the number of principal components used to represent the data (also see
Fig.~\varExplained). Several patterns were revealed by the analysis. First, in
general (i.e., across experimental conditions), decoding accuracy tends to
improve as the number of components are increased. However, decoding accuracy
peaked at higher levels for experimental conditions that exposed participants
to cognitively richer stimuli (Fig.~\ref{fig:inflection}D). The peak decoding
accuracy was highest for the ``intact'' condition (versus paragraph: $t(99) =
35.205, p < 0.001)$; versus word: $t(99) = 43.172, p < 0.001)$; versus rest:
$t(99) = 81.361, p < 0.001)$), next highest for the ``paragraph'' condition
(versus word: $t(99) = 6.243, p < 0.001)$; versus rest: $t(99) = 50.748, p <
0.001)$), and next highest for the ``word'' condition (versus rest: $t(99) =
48.791, p < 0.001)$). This ordering implies that cognitively richer conditions
evoke more stable brain activity patterns across people.

The cognitively richer conditions also displayed steeper initial slopes. For
example, the intact condition decoders reached an arbitrarily chosen threshold
of 5\% accuracy using fewer components than the paragraph condition decoders
($t(99) = -7.429, p < 0.001)$) or word condition decoders ($t(99) = -7.300, p <
0.001)$), and decoding accuracy never exceeded 5\% for the rest condition. This
suggests that brain activity patterns evoked by cognitively richer conditions
are more compressible, such that representing the data using the same number of
principal components provides more information to the temporal decoders
(Figs.~\ref{fig:inflection}B, C). Taken together, as shown in
Figure~\ref{fig:inflection}E, we found that brain activity patterns evoked by
cognitively richer conditions tended to be both more informative (i.e.,
associated with higher decoding accuracies) \textit{and} more compressible
(i.e., requiring fewer components to achieve the 5\% accuracy threshold).

@@

\label{fig:inflection-quarters}

@@

If informativeness (to the temporal decoders) and compressibility vary with the
cognitive richness of the stimulus, might these measures also vary over time
\textit{within} a given condition? For example, participants in the intact
condition might process the ongoing story more deeply later on in the story
(compared with earlier in the story) given the additional narrative background
and context they had been exposed to by that point. To examine this
possibility, we divided each condition into four successive time segments. We
computed decoding curves (Fig.~\ref{fig:inflection-quarters}A) and the numbers
of components required to achieve 5\% decoding accuracy
(Fig.~\ref{fig:inflection-quarters}B) for each segment and condition. We found
that, in the two most cognitively rich conditions (intact and paragraph), both
decoding accuracy and compressibility, as reflected by the change in decoding
curves, increased with listening time (e.g., at the annotated reference point
of $k = 20$ components in Fig.~\ref{fig:inflection-quarters}C: intact: $t(99) =
7.915, p < 0.001)$; paragraph: $t(99) = 2.354, p = 0.021)$). These changes may
reflect an increase in comprehension or depth of processing with listening
time. In contrast, the decoding accuracy and compressibility \textit{decreased}
with listening time in the word condition ($t(99) = -10.747, p < 0.001)$) and
rest condition ($t(99) = -22.081, p < 0.001)$). This might reflect the
depletion of attentional resources in the less-engaging word and rest
conditions.

@@

We also wondered how informativeness and compressibility in the different
experimental conditions might vary across brain networks. We used a network
parcellation identified by \cite{YeoEtal11} to segment the brain into seven
distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Figs.~\ref{fig:networks}A--C):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy tended to
increase, particularly in the higer-order experimental conditions
(Fig.~\ref{fig:networks}D, E). This suggests that higher-order networks may
carry more content-relevant or stimulus-driven ``information.'' We found no
clear trends in the proportions of components required to achieve 5\% decoding
accuracy across networks or conditions (Fig.~\ref{fig:networks}F).

@@

\caption{\textbf{Neurosynth topic weightings by component.} We used a reverse
inference procedure (see \textit{Reverse inference}) to estimate the
correspondences between brain images and a set of 80 topics derived from the
Neurosynth database of neuroimaging articles~\citep{RubiEtal17}. \textbf{A.
Topic correlations by component.} For each of the top five highest-weighted
principal components (columns) derived from each experimental condition
(colors), the radar plots display the correlations between the component images
and the per-topic images derived for each topic (topic identities are denoted
by the blue dots; a legend for the topic labels is in the lower right of the
figure). An annotated list of the top-weighted topics for each component and
condition may be found in Figure~\topTerms~and the top-weighted terms for each
topic may be found in Table~\topics. \textbf{B. Topic correlations averaged
across components.} The radar plot is in the same format as the plots in Panel
A, but here we display the per-condition average correlations across the top
five components (for each condition) reflected in Panel A. \textbf{C. Component
images.} Each plot diplays a right sagittal view of a glass brain projection of
the top five principal components (columns) for each experimental conditions
(rows). Additional projections for each component may be found in
Figure~\componentBrains.}

@@

In addition to examining different networks in isolation, we wondered about the
general structure of the full-brain (i.e., potentially multi-network) activity
patterns reflected by different principal components across different
experimental conditions. As shown in Figure~\ref{fig:neurosynth-pca}, we used
Neurosynth~\citep{RubiEtal17} to identify, for each component, the associations
with each of 80 themes (see \textit{Reverse inference}).  In general, the first
principal components across all of the experimental conditions tended to weight
most heavily on themes related to cognitive control, memory,
language processing, attention, and perception.

To elucidate....\textbf{JRM STOPPED HERE}

We noticed (by
inspection) several common themes across the sets of terms associated with each
component and condition. Memory-related components included terms like ``middle
temporal,'' ``memory retrieval,'' and ``working memory.'' Sensory processing
related components included terms like ``primary visual,'' ``auditory cortex,''
``v1,'' and so on. Other components were associated with sensory integration
(e.g., ``thalamic,'' ``cingulate,'' etc.), sentence comprehension (e.g.,
``sentence,'' ``comprehension''), emotion and valence (e.g., ``amygdala,''
``valance''), or the default mode network (e.g., ``default mode''). The
components we identified were relatively stable across story segments
(Fig.~\synthThirds).

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.9\textwidth]{figs/components_neurosynth_summary}

\caption{\textbf{Summary of functions associated with top-weighted components
by condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and several core cognitive function classifications is
displayed in the lower right of the figure. Table~\topics~provides a list of
each topic's top-weighted terms, along with each topic's manually labeled
cognitive classification. A full list of the topics most highly associated with
each component may be found in Figure~\topTerms. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heatmaps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heatmaps display the correlations between the brain maps
(Fig.~\componentBrains) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition components, by
component.} Each sub-panels network plot summarizes the pattern of correlations
between the $n$\textsuperscript{th} top-weighted principal components
(sub-panel) for each experimental condition (gray-outlined dots). The line
thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heatmaps in Panel C). \textbf{E. Change in
weights by condition.} The bar heights reflect the slopes of regression lines,
fit separately to the top five components from each condition, between the
ChatGPT-derived ``rank'' of each cognitive classification and the correlations
between the component and topic maps associated with cognitive processes at the
given rank (see \textit{Ranking cognitive processes}).  Also see
Fig.~\neurosynthFull~for additional information.}
